|

### 2023-05-11 paper/03-methods.Rmd
fixed effects table

In order to apply strain rate to measure NFL pass rusher effectiveness, we make modifications to how this concept is traditionally defined.

Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of pass rusher $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.

*   The distance between pass rusher $j$ and the quarterback at frame $t$ during play $i$ is 
\begin{equation}
f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
\end{equation}
*   The velocity at which pass ruher $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
\begin{equation}
f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}.
\end{equation}
*   The STRAIN for pass rusher $j$ at frame $t$ during play $i$ is 
\begin{equation}
\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)}.
\end{equation}
Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters (STRAIN) for the remainder of this manuscript.

@@

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers linked to a rusher, for simplicity we only consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\beta$ for additional information $X_{j,i}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{r[ij]} + B_{b[ij]} + Q_{q[i]} + X_{ij} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play, the pass rusher position, the nearest pass blocker's position, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

>>>>

In order to apply strain rate to measure NFL pass rusher effectiveness, we make modifications to how this concept is traditionally defined.
Let $(x_{ijt}, y_{ijt})$ be the $(x, y)$ location on the field of pass rusher $j = 1, \cdots, J$ at frame $t = 1, \cdots, T_i$ for play $i = 1, \cdots, n$; and $(x^{QB}_{it}, y^{QB}_{it})$ be the $(x, y)$ location of the quarterback at frame $t$ during play $i$.

*   The distance between pass rusher $j$ and the quarterback at frame $t$ during play $i$ is 
\begin{equation}
f_{d_{ij}}(t) = \sqrt{(x_{ijt} - x^{QB}_{it})^2 + (y_{ijt} - y^{QB}_{it})^2}.
\end{equation}
*   The velocity at which pass ruher $j$ is moving towards the quarterback at frame $t$ during play $i$ is 
\begin{equation}
f'_{d_{ij}}(t) = \frac{df_{d_{ij}}(t)}{dt}.
\end{equation}
*   The STRAIN for pass rusher $j$ at frame $t$ during play $i$ is 
\begin{equation}
\text{STRAIN}_{ij}(t) = \frac{-f'_{d_{ij}}(t)}{f_{d_{ij}}(t)}.
\end{equation}

Note that to distinguish our metric from strain and strain rate in materials science, we write it in capital letters (STRAIN) for the remainder of this manuscript.

@@

To this end, we fit a multilevel model to evaluate pass rushers' impact on the average STRAIN observed in a play, while accounting for their pass blocker, QB, and play-level information.
We identify the pass blocker linked with the pass rusher of interest using the data provided by PFF.
Since there can be multiple blockers linked to a rusher, for simplicity we only consider the nearest blocker positioned to the pass rusher at the start of the play.
We use random intercepts for the three player groups: (1) pass rushers as $R$, (2) nearest pass blockers as $B$, and (3) QBs as $Q$.
The coefficients $\beta$ for additional information $X_{ij}$ about pass rusher $j$ in play $i$ are estimated as fixed-effects. 
Our model for the average STRAIN by pass rusher $j$ on play $i$ is as follows.
\begin{equation}
\begin{aligned}
\overline{\text{STRAIN}}_{ij} &\sim N(R_{r[ij]} + B_{b[ij]} + Q_{q[i]} + X_{ij} \cdot \beta, \sigma^2), \text{ for } i = 1, \dots, n \text{ plays} \\
R_{r} &\sim N(\mu_R, \sigma^2_R), \text{ for } r = 1, \cdots, \text{ \# of pass rushers}, \\
B_{b} &\sim N(\mu_B, \sigma^2_B), \text{ for } b = 1, \cdots, \text{ \# of pass blockers}, \\
Q_{q} &\sim N(\mu_Q, \sigma^2_Q), \text{ for } q = 1, \cdots, \text{ \# of QBs}.
\end{aligned}
\end{equation}

We consider a normal distribution to obtain the random intercepts for each player toward their respective position group means.
This provides us with average individual effects on STRAIN for each player while applying shrinkage, which is a useful property in this problem's context since we do not observe the same number of plays for each player.
With regards to the fixed-effects about pass rusher $j$ in play $i$, we include the length of the play (in terms of elapsed time between ball snap and a QB event), the positions of pass rusher and nearest blocker, and the number of pass blockers assigned to the pass rusher based on the PFF data.
Table \ref{tab:pos} shows our positional categorization for the pass rush and pass block roles.
We implement the model using penalized likelihood via the `lme4` package in `R` [@bateslme4; @R2023Language].

### 2023-05-11 paper/04-results.Rmd
fixed effects table

[^unsuccess]: https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb

@@

Further, Figure \ref{fig:fig_outcome_curves} shows the positional STRAIN curves by whether a play's outcome is a hit, sack, hurry, or none of those. 
We clearly see that players tend to generate more STRAIN when a play ends in a hit, sack, or hurry, compared to no outcome.
Within the three pressure metrics, it is not surprised that a sack corresponds to the highest amount of STRAIN.
[Can we say anything about why NT has highest curve when there's a sack?]

@@

\begin{table}
\caption{Fixed-effects coefficients estimates for the multilevel model. \label{tab:fixedeff}}
\centering
\begin{tabular}{rrrrrr}
  \hline
 & estimate & se & df & $t$-statistic & $p$-value \\ 
  \hline
Intercept & 0.355 & 0.010 & 332.288 & 35.688 & 0.000 \\ 
  play\_len & -0.004 & 0.000 & 30653.793 & -38.214 & 0.000 \\ 
  rush\_posDT & -0.081 & 0.008 & 369.713 & -10.340 & 0.000 \\ 
  rush\_posinteriorLB & 0.058 & 0.011 & 770.441 & 5.332 & 0.000 \\ 
  rush\_posNT & -0.106 & 0.012 & 355.999 & -9.173 & 0.000 \\ 
  rush\_posOLB & 0.062 & 0.008 & 388.506 & 7.695 & 0.000 \\ 
  rush\_possecondary & 0.145 & 0.012 & 1734.779 & 12.328 & 0.000 \\ 
  block\_posG & 0.011 & 0.006 & 123.672 & 1.867 & 0.064 \\ 
  block\_posother & 0.010 & 0.008 & 500.451 & 1.261 & 0.208 \\ 
  block\_posT & 0.023 & 0.006 & 161.556 & 3.625 & 0.000 \\ 
  n\_blockers & -0.064 & 0.003 & 30949.514 & -24.788 & 0.000 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\caption{ICC. \label{tab:icc}}
\centering
\begin{tabular}{rlr}
\hline
Group & ICC \\ 
\hline
Pass rusher & 0.057 \\ 
Pass blocker & 0.012 \\ 
Quarterback & 0.023 \\ 
Residual & 0.907 \\ 
\hline
\end{tabular}
\end{table}

>>>>

[^unsuccess]: A real-game footage of this play is available at https://www.nfl.com/videos/teddy-bridgewater-courtland-sutton-barely-miss-out-on-chance-at-62-yard-td-bomb

@@

Further, Figure \ref{fig:fig_outcome_curves} shows the positional STRAIN curves by whether a play's outcome is a hit, sack, hurry, or none of those. 
We clearly see that players tend to generate more STRAIN when a play ends in a hit, sack, or hurry, compared to no outcome.
Within the three pressure metrics, it is not surprised that a sack corresponds to the highest amount of STRAIN, followed by a hit and hurry.

@@

<!-- reference levels: DE (rush), T (block) -->

\begin{table}
\caption{Fixed-effects coefficients estimates for the multilevel model for average STRAIN. Note that the reference pass rusher level is defensive end (DE), denoted PR:DE; and the reference pass blocker level is tackle (T), denoted PB:T. \label{tab:fixedeff}}
\centering
\begin{tabular}{lrrrr}
\hline
& estimate & se & $t$-statistic & $p$-value \\ 
\hline
Intercept & 0.378 & 0.009 & 43.310 & 0.000 \\ 
Length of play & -0.004 & 0.000 & -38.214 & 0.000 \\
Number of blockers & -0.064 & 0.003 & -24.788 & 0.000 \\
$I_{\text{\{PR:DT\}}}$ & -0.081 & 0.008 & -10.340 & 0.000 \\
$I_{\text{\{PR:interior\}}}$ & 0.058 & 0.011 & 5.332 & 0.000 \\
$I_{\text{\{PR:NT\}}}$ & -0.106 & 0.012 & -9.174 & 0.000 \\
$I_{\text{\{PR:OLB\}}}$ & 0.062 & 0.008 & 7.695 & 0.000 \\
$I_{\text{\{PR:secondary\}}}$ & 0.145 & 0.012 & 12.328 & 0.000 \\
$I_{\text{\{PB:C\}}}$ & -0.023 & 0.006 & -3.625 & 0.000 \\
$I_{\text{\{PB:G\}}}$ & -0.011 & 0.005 & -2.290 & 0.023 \\ 
$I_{\text{\{PB:other\}}}$ & -0.013 & 0.007 & -1.911 & 0.056 \\
\hline
\end{tabular}
\end{table}

<!-- (Intercept) & 0.378 & 0.009 & 43.310 & 0.000 \\  -->
<!--   play\_len & -0.004 & 0.000 & -38.214 & 0.000 \\  -->
<!--   rush\_posDT & -0.081 & 0.008 & -10.340 & 0.000 \\  -->
<!--   rush\_posinteriorLB & 0.058 & 0.011 & 5.332 & 0.000 \\  -->
<!--   rush\_posNT & -0.106 & 0.012 & -9.174 & 0.000 \\  -->
<!--   rush\_posOLB & 0.062 & 0.008 & 7.695 & 0.000 \\  -->
<!--   rush\_possecondary & 0.145 & 0.012 & 12.328 & 0.000 \\  -->
<!--   block\_posC & -0.023 & 0.006 & -3.625 & 0.000 \\  -->
<!--   block\_posG & -0.011 & 0.005 & -2.290 & 0.023 \\  -->
<!--   block\_posother & -0.013 & 0.007 & -1.911 & 0.056 \\  -->
<!--   n\_blockers & -0.064 & 0.003 & -24.788 & 0.000 \\  -->

\begin{table}[ht]
\caption{Intraclass correlation coefficients for the multilevel model for average STRAIN. \label{tab:icc}}
\centering
\begin{tabular}{lr}
\hline
Group & ICC \\ 
\hline
Pass rusher & 0.057 \\ 
Pass blocker & 0.012 \\ 
Quarterback & 0.023 \\ 
Residual & 0.907 \\ 
\hline
\end{tabular}
\end{table}

### 2023-05-11 paper/05-discussion.Rmd
fixed effects table

@@

(Stuff from BDB draft)

>>>>

rapm

@@

from bdb submission

As for future directions, STRAIN could be utilized to collectively evaluate the performance of a pass-rushing unit by considering the total amount of STRAIN during a play, which could, in turn, be used to examine the effectiveness of different types of blitzes. Moreover, while we focused solely on pass-rushers, STRAIN can be easily extended to the assessment of pass-blockers as a unit by looking at quantities such as the total STRAIN or the maximum STRAIN per frame aggregated across the entire play. It is also possible to apply STRAIN to assess individual offensive pass-blockers, provided that a method of matching blockers to rushers was developed.

Our metric, however, is subject to several limitations. As currently devised, STRAIN cannot distinguish between a pass-rusher who has to face a single blocker versus a double team. This problem could be addressed, at least somewhat, by having a method of matching pass-rushers to blockers as previously mentioned. Additionally, there are a few situations such as quarterback out-of-pocket scrambling where STRAIN does not appear to be useful. If a pass-rusher is on the left side and the quarterback scrambles out to the left, the STRAIN for that pass-rusher can drop considerably as the distance between the quarterback and pass-rusher increases, which the defender should not necessarily be negatively affected by in a measure of pass-rushing effectiveness. -->

from bdb draft
